Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-24 _tex/index.tex
Built site for gh-pages

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's structural and measurement model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role and impact of comparison algorithms and the number of comparisons
on the method's reliability and validity. The following sections will
discuss each of these issues in detail, followed by the introduction of
a theoretical model and its statistical translation, which aim to
address all three concerns simultaneously.

\section{The three critical issues in CJ
literature}\label{sec-theory-issues}

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model.
This model specifies how latent variables are estimated from observed
variables \citep{Everitt_et_al_2010}. In CJ, multiple judges engage in
several rounds of pairwise comparisons to assess the relative
manifestation of a trait between two stimuli. Each comparison generates
a dichotomous outcome, indicating which stimulus is perceived to exhibit
a higher degree of the trait. The BTL model then uses these observed
outcomes to estimate scores that represent the latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, researchers often use these BTL-generated scores or their
transformations separately, conducting further analyses or hypothesis
testing. The CJ literature highlights how these scores have been
employed to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores for separate analyses and hypothesis testing. A key consideration
is that BTL-generated scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty when conducting separate
analyses and test can inflate their precision and statistical power.
This, in turn, increases the risk of committing a type I error, which is
when a null hypothesis is incorrectly rejected \citep{McElreath_2020}.
To mitigate this risk, principles from Structural Equation Modeling
(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting
these analyses within a structural model that accounts for both the
scores and their uncertainties, rather than treating them separately.
Thus, an integrated approach combining CJ's structural and measurement
models can offer significant advantages.

>>>>

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's structural and measurement model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role and impact of comparison algorithms on the method's reliability and
validity. The following sections will discuss each of these issues in
detail, followed by the introduction of a theoretical model and its
statistical translation, which aim to address all three concerns
simultaneously.

\section{Three critical issues in CJ
literature}\label{sec-theory-issues}

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
\citep{Andrich_1978, Bramley_2008}. This model specifies how latent
variables are estimated from manifest variables
\citep{Everitt_et_al_2010}. In CJ, multiple judges engage in several
rounds of pairwise comparisons to assess the relative manifestation of a
trait between two stimuli. Each comparison generates a dichotomous
outcome, indicating which stimulus is perceived to exhibit a higher
degree of the trait. The BTL model then uses these observed outcomes to
estimate scores that represent the latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, researchers often use these BTL-generated scores or their
transformations separately, conducting further analyses or hypothesis
testing. The CJ literature shows how these scores have been employed to
identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores for separate analyses and tests. A key consideration is that
BTL-generated scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting separate analyses
and tests can inflate their precision and statistical power. This, in
turn, increases the risk of committing a type I error
\citep{McElreath_2020}, which is when a null hypothesis is incorrectly
rejected \citep{Everitt_et_al_2010}. To mitigate this risk, principles
from Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response Theory
(IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
conducting these analyses and tests within a structural model that
accounts for both the scores and their uncertainties, rather than
treating them separately. Thus, an integrated approach combining CJ's
structural and measurement models can offer significant advantages.
